---
title: "dplyr_dora_ranilovic"
author: "Dora Ranilovic"
date: "November 4, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Warm-up

### Make sure you've installed {dplyr} >= 0.7 and {prenoms} package

```{r eval=FALSE, echo=FALSE}

##installed
c("dplyr","prenoms") %in% installed.packages()

```


### Load here `{dplyr}`, `{prenoms}` and any other needed package

```{r}
library(dplyr)
library(prenoms)
library(tidyr)
library(ggplot2)
library(stringr)


```

### Import

#### prenomsdataset

Using `data(prenoms)` load `prenoms` dataset from  `{prenoms}` package.

```{r}
data("prenoms")
```

What kind of object is `prenoms` ? 

```{r}
class(prenoms)
#'prenoms' is a dataframe
```

Explore the database using the '5-functions-to-always-run-on-a-database'

```{r, eval=FALSE}
dim(prenoms)
names(prenoms)
head(prenoms)
View (prenoms)
summary(prenoms)

```

Using `glimpse`, have a look at `prenoms`'s structure.

```{r}
glimpse(prenoms)

```

#### Regions, departements and surfaces

Load the "dpt_data_modif.csv" dataset from IGN (French public state administrative establishment founded in 1940[1] to produce and maintain geographical information for France and its overseas departments and territories) using the appropriate function. Data have been prepared for you: the surface of departement has been calculated and spatial data removed.

```{r}
dept_data_modif <- read.csv("data/dpt_data_modif.csv",encoding = "latin1")
```


#### Elementary and college schools

We also fetched for you on [data.gouv.fr](https://www.data.gouv.fr/fr/datasets/adresse-et-geolocalisation-des-etablissements-denseignement-du-premier-et-second-degres/#_) the addresses of "primary and secondary schools, the administrative structures of the Ministry of National Education. Public and private sectors."

1. Data preprocessing 
    + Import the csv file : "DEPP-etab-1D2D.csv" and name it "depp_orig"
        + Encoding is `"latin1"`
    + Transform zip code ("code_postal_uai") into 5 characters with zeros
    + Extract department numbers ("dpt") starting from column "code_postal_uai"
    + Save the modifications into "depp_modif.csv"

```{r}
depp_orig <- read.csv2("data/DEPP-etab-1D2D.csv", encoding = "latin1")

#View(depp_orig)
code_postal_modif <- str_pad((as.character(depp_orig$code_postal_uai)),5,pad="0")

dpt <- substr(code_postal_modif,start=1,stop=2)

depp_modif <- depp_orig %>% mutate(code_postal_uai=code_postal_modif,dpt=dpt)

#write.csv(depp_modif, "data/depp_modif.csv")

```

2. Read the pre-processed "depp_modif.csv" file

```{r}
depp_modif <- read.csv("data/depp_modif.csv")
```


#### Facts observed by the police services and national gendarmerie units by department

We also gathered data from [data.gouv.fr](https://www.data.gouv.fr/fr/datasets/faits-constates-par-index-par-departement-et-par-annee/#_) concerning "all the facts observed by the police services and national gendarmerie units by department from 1996 to 2011"

1. Data preprocessing 
    - Import Excel sheet "2010" from "faitsconstatespardepartementde2002-a-2011.xls" file
        + _beware of the original formatting_
    - Copy it into "faits_2010_modif" in order to make some modifications:
        + Delete Excel calculations:
            + `Tout_département`, `Tout_index`
        + Transform in long format using `gather`
            + 4 columns : Index, Libellé, dpt, nombre
        + save the dataframe into a csv file "faits_2010_modif.csv"

```{r}
faits_2010_modif <- readxl::read_xls("data/faitsconstatespardepartementde2002-a-2011.xls", sheet="2010",range = "A3:CV107")
#View(faits_2010_modif)

faits_2010_modif <- faits_2010_modif %>% select(-c("Tout département","Service à compétence national")) %>% 
  filter(!is.na(Index) )%>%
  gather(dpt, nombre,"01":"95")

#write.csv(depp_modif, "data/faits_2010_modif.csv")
```

2. Read preprocessed file "faits_2010_modif.csv"

```{r}
faits_2010_modif <- read.csv("data/faits_2010_modif.csv", encoding = "latin1")
```


## Analyses

Some assumptions to do the exercise:

- every child born in a department stays in that department until the end of college
- every child between 11 and 14 years old is in a college
- the number of colleges is constant between 2010 and 2016
- College "à ouvrir" (i.e. "to be open") do not have children. Others have.

### Filter datasets to Metropolitan France

Datasets to be filtered: `prenoms`, `depp_modif`, `faits_2010_modif`, `dpt_data_modif`

- Department named "2A" and "2B" should be merged to "20"
- We only work with data in Metropolitan France, which means for "dpt" between `01` and `95` included. Others needs to be filtered.

```{r}

prenoms <- prenoms %>% 
  filter(dpt %in% str_pad(as.character((1:95)),2, pad="0")) 

depp_modif <- depp_modif %>% 
  mutate(dpt=as.character(dpt)) %>% 
  mutate(dpt = str_pad(dpt,side="left",width=2, pad="0")) %>% 
  filter(dpt %in% str_pad(as.character((1:95)),side="left",2, pad="0")) %>% 
  filter(etat_etablissement==1) 
  

faits_2010_modif <-  faits_2010_modif %>% 
  mutate(dpt=replace(dpt,dpt %in% c("2A","2B"),"20"))%>% 
  filter(dpt %in% str_pad(as.character((1:95)),2, pad="0")) 


dept_data_modif <- dept_data_modif %>% 
  mutate(CODE_DEPT = as.character(CODE_DEPT)) %>% 
  mutate(CODE_DEPT=replace(CODE_DEPT,CODE_DEPT %in% c("2A","2B"),"20"))  %>% 
  filter(CODE_DEPT %in% str_pad(as.character((1:95)),2, pad="0"))

```


### National average number of children per college in 2010 ?

```{r}
##total number of clildren aged 11-14 in 2010
coll_children_total <- prenoms %>% filter(year %in% (1996:1999)) %>% summarise(sum(n))



##total number of colleges
college_total <- depp_modif %>% filter(nature_uai %in% c(340,350,352)) %>% summarise(n())


##compute avg children per college
coll_children_total/college_total




```

### Average number of children per college in 2010 in each department?

- Arrange departments according to the calculated average in descending order

```{r}
##total number of clildren aged 11-14 in 2010 per dept
college_children_dept <- prenoms %>% 
  filter(year %in% (1996:1999)) %>% 
  group_by(dpt) %>% 
  summarise(children=sum(n)) %>% 
  ungroup()
college_children_dept

##total number of colleges
college_dept <- depp_modif %>%
  filter(nature_uai %in% c(340,350,352)) %>% 
  group_by(dpt) %>% 
  summarise(n_college=n()) %>% 
  ungroup()
college_dept

##compute avg children per college

college_children_dept %>% inner_join(college_dept,"dpt") %>% 
  mutate (average_children_per_college=children/n_college) %>% 
  arrange(desc(average_children_per_college))



```

### Number of Facts observed by the police services in 2010 per department ? 

```{r}
faits_2010_modif %>% group_by(dpt) %>% summarise(Facts=sum(nombre)) %>% arrange(dpt)

```

### Number of children born, number of colleges and facts related by the police services per department in 2010 ?

- Group all information in the same table
- Arrange by descending order of children, schools and facts

```{r}
faits_2010_modif %>% 
  group_by(dpt) %>% 
  summarise(Facts=sum(nombre))  %>% 
  left_join(prenoms %>% 
               filter(year==2010) %>% 
               group_by(dpt) %>% 
               summarise(born=sum(n))) %>% 
               replace(.,is.na(.),0) %>% 
  left_join(depp_modif %>% 
               filter(nature_uai%in% c(340,350,352)) %>% 
               group_by(dpt) %>%
               summarise(n_college=n())) %>% 
               mutate(n_college = ifelse(is.na(n_college), 0, n_college)) %>% 
  select(dpt,born,n_college,Facts) %>% 
  arrange(desc(born),desc(n_college),desc(Facts))


```

### Number of children born, number of colleges and facts related by the police services per km² in 2010 by department?

```{r}


stats_per_km <- 
  faits_2010_modif %>%
  group_by(dpt) %>% 
  summarise(Facts=sum(nombre)) %>%
  mutate(Facts = ifelse(is.na(Facts), 0, Facts)) %>%    
  left_join(dept_data_modif %>% 
               mutate(dpt=CODE_DEPT,surface=surface_m/1000000) %>% 
               select(dpt,surface) %>% 
               group_by(dpt) %>% 
               summarise(surface=sum(surface)), by="dpt")  %>% 
  replace(.,is.na(.),0)   %>% 
  mutate(facts_per_km = Facts/surface)  %>% 
  left_join(prenoms %>% 
               filter(year==2010) %>% 
               group_by(dpt) %>% 
               summarise(born=sum(n)),by="dpt") %>% 
  replace(.,is.na(.),0)  %>% 
  mutate(born_per_km= born/surface)  %>%  
  left_join(depp_modif %>% 
               mutate(dpt=replace(dpt,dpt %in% c("2A","2B"),"20")) %>% 
               filter(nature_uai %in% c(340,350,352)) %>% 
               group_by(dpt) %>%
               summarise(n_college=n()),by="dpt") %>% 
  replace(.,is.na(.),0)  %>% 
  mutate(schools_per_km=n_college/surface) %>%  
  select(dpt,born_per_km,schools_per_km,facts_per_km) %>% 
  arrange(desc(born_per_km),desc(schools_per_km),desc(facts_per_km))



```

### Is there a correlation between the number of birth and the number of facts related by the police per km² in 2010 ?

There seems to be a strong correlation.

```{r}
ggplot(data=stats_per_km,aes(x=born_per_km,y=facts_per_km))+geom_point() 
#+coord_cartesian(xlim=c(0,250),ylim = c(0,2000))


cor(stats_per_km$born_per_km, stats_per_km$facts_per_km,method = "pearson",use = "complete.obs")


```

### What is the mean regional density (in number/km²) of the 15 most given first names in France ?

- Filter the 15 most given first names in France
- Create a wide table with the department as observations and the 15 most given names in columns (as variables): the count is at the row-column intersection 
- merge with the surface department infos
- Compute the region surface and the sum of names by region

```{r}

top_names <- prenoms %>% group_by(name) %>% 
  summarise(n=sum(n)) %>% 
  arrange(desc(n)) %>% 
  top_n(15) 


#top_names_wide <-
  prenoms %>% filter(name %in% top_names$name) %>% 
  group_by(dpt,name) %>% 
  summarise(n=sum(n)) %>% 
  spread(key = name,value=n) %>% 
  inner_join(dept_data_modif %>% 
               select(CODE_DEPT,surface_m) %>% 
               mutate(dpt=CODE_DEPT), by="dpt") %>% 
  select(-CODE_DEPT) %>% 
  mutate(surface = surface_m/1000000) %>% 
  mutate_if(is.numeric, funs(density=./surface)) %>% 
  select(c(1, 19:33))




```

#### Bonus question : map the mean regional density (in number/km²) of the 15 most given first names in France 

- Use the "department" shapefile to cross information and map data

```{r}






